---
title: "public bike sharing data"
author: "Diana Coman Schmid"
date: "3 Dezember 2016"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_chunk$set(include = FALSE)
```

Install and load required R packages:
```{r,eval=FALSE,echo=TRUE, include=TRUE}
# Check if packages are installed (install them if they are not), then load them into the R session.

ipak <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg))
    install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)
}


packages <- c("ggplot2", "ggmap","rgdal","rgeos","maptools","dplyr","tidyr","tmap","maps","caret",
              "rpart","rpart.plot","data.table","timeDate","lubridate","randomForest","MASS","ranger","igraph")

#ipak(packages)


lapply(packages, library,character.only=TRUE)

```

```{r,eval=TRUE,echo=FALSE, include=FALSE}
# Check if packages are installed (install them if they are not), then load them into the R session.

# ipak <- function(pkg){
#   new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
#   if (length(new.pkg))
#     install.packages(new.pkg, dependencies = TRUE)
#   sapply(pkg, require, character.only = TRUE)
# }


packages <- c("ggplot2", "ggmap","rgdal","rgeos","maptools","dplyr","tidyr","tmap","maps","caret",
              "rpart","rpart.plot","data.table","timeDate","lubridate","randomForest","MASS","ranger","igraph")

#ipak(packages)


lapply(packages, library,character.only=TRUE)

```

# **Import the data**
* Parameters as described at data source (https://www.citibikenyc.com/system-data)
    + Trip Duration (seconds)
    + Start Time and Date
    + Stop Time and Date
    + Start Station Name
    + End Station Name
    + Station ID
    + Station Lat/Long
    + Bike ID
    + User Type (Customer = 24-hour pass or 7-day pass user; Subscriber = Annual Member)
    + Gender (Zero=unknown; 1=male; 2=female)
    + Year of Birth

*This data has been processed to remove trips that are taken by staff as they service 
    and inspect the system, trips that are taken to/from any of our "test" stations (which
    we were using more in June and July 2013), and any trips that were below 60 seconds in 
    length (potentially false starts or users trying to re-dock a bike to ensure it's secure).*

```{r,eval=TRUE,echo=TRUE,include=TRUE}

# import speed comparison
# system.time(read.csv("E://eFeel//201602-citibike-tripdata.csv",header=TRUE))
# system.time(fread("E://eFeel//201602-citibike-tripdata.csv",header=TRUE))


ef <- fread("C://Users//Diana//Desktop//eFeel//201602-citibike-tripdata.csv",header=TRUE)

#/// on Mac
# temp <- tempfile()
# download.file("https://s3.amazonaws.com/tripdata/201602-citibike-tripdata.zip",temp, method="curl")
# unzip(temp, file = "201602-citibike-tripdata.csv")
# ef <- fread("201602-citibike-tripdata.csv",header=TRUE)
# unlink(temp)

str(ef)

```

# **Data preprocessing**

```{r,eval=TRUE,echo=TRUE}

# remove spaces in column names
colnames(ef) <- gsub(" ","_",colnames(ef))

# set to factor/numeric where relevant
ef$tripduration <- as.numeric(ef$tripduration)
ef$start_station_id <- as.factor(ef$start_station_id)
ef$start_station_name <- gsub(" ","_",ef$start_station_name)
ef$start_station_name <- as.factor(ef$start_station_name)
ef$start_station_latitude <- as.numeric(ef$start_station_latitude)
ef$start_station_longitude <- as.numeric(ef$start_station_longitude)
ef$end_station_id <- as.factor(ef$end_station_id)
ef$end_station_name <- gsub(" ","_",ef$end_station_name)
ef$end_station_name <- as.factor(ef$end_station_name)
ef$end_station_latitude <- as.numeric(ef$end_station_latitude)
ef$end_station_longitude <- as.numeric(ef$end_station_longitude)
ef$bikeid <- as.factor(ef$bikeid)
ef$usertype <- as.factor(ef$usertype)
ef$birth_year <- as.numeric(ef$birth_year)
ef$gender <- as.factor(ef$gender)

# note the data class of fread()
class(ef)
ef.s <- as.data.frame(ef)

# filter out irrealistic values of parameters
# tripduration (e.g. 1811000 sec = 503 hours...) -- typing error? 

# summary(ef.s$tripduration)
# quantile(ef.s$tripduration, c(0.05,0.95))

par(mfrow=c(1,2))
boxplot(log2(ef.s$tripduration),outcol="gray",main="tripduration (sec.; log2)",cex.main=0.75)
abline(h=quantile(log2(ef.s$tripduration), c(0.05,0.95)),col="cornflowerblue")
mtext(c("q5","q95"),side=4,at = quantile(log2(ef.s$tripduration), c(0.05,0.95)),col="cornflowerblue")


plot(density(log2(ef.s$tripduration)),col="gray",main="tripduration (sec.; log2) distribution",cex.main=0.75)
abline(v=quantile(log2(ef.s$tripduration), c(0.05,0.95)),col="cornflowerblue")
mtext(c("q5","q95"),side = 1,at = quantile(log2(ef.s$tripduration), c(0.05,0.95)),col="cornflowerblue")


tripd.s <- which(ef.s$tripduration >= quantile(ef.s$tripduration,0.05) &
                   ef.s$tripduration <= quantile(ef.s$tripduration,0.95))
#str(tripd.s)  
#summary(ef.s$tripduration[tripd.s])


# birth year (e.g. people born in 1885 are 131 in 2016 :) -- typing error 1885 instead of 1985?
# note the Qs
# quantile(ef.s$birth_year, c(0.005,0.995),na.rm = TRUE)


par(mfrow=c(1,2))
boxplot(ef.s$birth_year,outcol="gray",main="birth year",cex.main=0.75,na.rm=TRUE)
abline(h=quantile(ef.s$birth_year, c(0.005,0.995),na.rm=TRUE),col="cornflowerblue")
mtext(c("q0.5","q99"),side=4,at = quantile(ef.s$birth_year, c(0.005,0.995),na.rm=TRUE),col="cornflowerblue")


plot(density(ef.s$birth_year,na.rm=TRUE),main="birth year distribution",cex.main=0.75)
abline(v=quantile(ef.s$birth_year, c(0.005,0.995),na.rm=TRUE),col="cornflowerblue")
mtext(c("q0.5","q99"),side = 1,at = quantile(ef.s$birth_year, c(0.005,0.995),na.rm=TRUE),col="cornflowerblue")


byear.s <- which(ef.s$birth_year >= quantile(ef.s$birth_year,0.005,na.rm = TRUE) &
                   ef.s$birth_year <= quantile(ef.s$birth_year,0.995,na.rm = TRUE))


# define filter based on realistic tripduration and birth year values
sel <- which(ef.s$tripduration >= quantile(ef.s$tripduration,0.05) &
             ef.s$tripduration <= quantile(ef.s$tripduration,0.95) &
             ef.s$birth_year >= quantile(ef.s$birth_year,0.005,na.rm = TRUE) &
             ef.s$birth_year <= quantile(ef.s$birth_year,0.995,na.rm = TRUE))

#str(sel)


# subset data based on relevant parameter values
ef.sel <- ef.s[sel,]

# create time related variables 
# week day; working day//week-end; hours; periods (e.g. night, morning, noon, etc.)

ef.sel$timedate.start <-mdy_hms(ef.sel$starttime) 
ef.sel$hours.start <- hour(ef.sel$timedate.start)
#ef.sel$days.start <- wday(ef.sel$timedate.start)
ef.sel$days.start <- wday(ef.sel$timedate.start,label=TRUE)
ef.sel$weeks.start <- ceiling(day(ef.sel$timedate.start)/7)
ef.sel$workday.start <- as.numeric(isWeekday(ef.sel$timedate.start))
# ef.sel$pernum.start <- cut(ef.sel$hours.start, c(-Inf, 7, 10, 12, 17, Inf),
#                                  labels=c(0, 1, 2, 3, 4))
ef.sel$pernum.start <- cut(ef.sel$hours.start, c(-Inf, 7, 10, 12, 17, Inf),
                                 labels=c("night", "morning", "noon", "afternoon", "evening"))

ef.sel$age <- cut(ef.sel$birth_year, c(-Inf,1960,1980,Inf),
                            labels=c("old","mid_age","young"),
                              right=T)


# round time to nearest hour
ef.sel$nearestH.start <- floor_date(ef.sel$timedate.start,"hour")

str(ef.sel)
summary(ef.sel)

```

# Explore the data

```{r,eval=TRUE,echo=TRUE,include=TRUE}

# total number of stations (start)
length(levels(ef.sel$start_station_name))

# usertype
# there is no birth year info for Customers (obviously; see data description at source)
# the subsequent analyses ignore the Customers

table(ef.s$usertype)
table(ef.sel$usertype)


par(mfrow=c(1,1))
hist(ef.s$birth_year,main="birth year distribution for Customers and Subscribers",cex.main=0.75,xlab="year",col=rgb(0.1,0.1,0.1,0.25))
hist(ef.s[which(ef.sel$usertype == "Subscriber"),"birth_year"],main="birth year distribution for Subscribers",cex.main=0.75,xlab="year",col=rgb(1,1,0,0.25),add=TRUE)

yr.cust <- ef.s[which(ef.s$usertype == "Customer"),"birth_year"]
str(yr.cust)

# gender

bpg <- barplot(table(ef.sel$gender),main="number of women and men",cex.main=0.75,xaxt="n")
axis(1,at = bpg,labels = c("na","men","women"),las=2)


# number of bikes in total
# length(unique(ef.s$bikeid))
length(unique(ef.sel$bikeid))

# time span in days
difftime(max(ef.sel$timedate.start),min(ef.sel$timedate.start))

# time (week days, time periods)

par(mfrow=c(1,3))
barplot(table(ef.sel$days.start),main="count week days",cex.main=0.75,las=2)
bpw <- barplot(table(ef.sel$workday.start),main="count of week-end days // work days",cex.main=0.75,xaxt="n")
axis(1,at = bpw,labels = c("week-end","work day"),las=2)
bpp <- barplot(table(ef.sel$pernum.start),main="count of time period ",cex.main=0.75,las=2)#,xaxt="n")

# bike demand per station, hour, week day, working day, gender, age
# bike count per age and hour;

bikeage.fr.start <- aggregate(bikeid ~ hours.start+age, data = ef.sel, FUN = function(x){NROW(unique(x))})
head(bikeage.fr.start)


peak.bikeage.fr.start <- bikeage.fr.start[which(bikeage.fr.start$bikeid >= 
                                              quantile(bikeage.fr.start$bikeid,0.95)),]

high.bikeage.fr.start <- bikeage.fr.start[which(bikeage.fr.start$bikeid < 
                                              quantile(bikeage.fr.start$bikeid,0.95) &
                                              bikeage.fr.start$bikeid >= 
                                              quantile(bikeage.fr.start$bikeid,0.75)),]


par(mfrow=c(1,1))
ggplot(data=bikeage.fr.start, aes(x=hours.start,y=bikeid,group=1)) + 
  geom_line(col="gray")+
  geom_point(col="gray")+
  geom_point(data=peak.bikeage.fr.start, colour="red",size=3) +
  geom_point(data=high.bikeage.fr.start, colour="orange",size=2) +
  geom_text(data=peak.bikeage.fr.start, label=peak.bikeage.fr.start$hours.start, 
            vjust= -1,col="gray",size=3) +
  facet_grid(~ age) +
  ylab("bike count")

# number of bikes per hour per day
bikeh.fr.start <- aggregate(bikeid ~ hours.start+days.start, data = ef.sel, FUN = function(x){NROW(unique(x))})
head(bikeh.fr.start)


peak.bikeh.fr.start <- bikeh.fr.start[which(bikeh.fr.start$bikeid >= 
                                              quantile(bikeh.fr.start$bikeid,0.95)),]

high.bikeh.fr.start <- bikeh.fr.start[which(bikeh.fr.start$bikeid < 
                                              quantile(bikeh.fr.start$bikeid,0.95) &
                                              bikeh.fr.start$bikeid >= 
                                              quantile(bikeh.fr.start$bikeid,0.75)),]


par(mfrow=c(1,1))
ggplot(data=bikeh.fr.start, aes(x=hours.start,y=bikeid,group=1)) + 
  geom_line(col="gray")+
  geom_point(col="gray")+
  geom_point(data=peak.bikeh.fr.start, colour="red",size=3) +
  geom_point(data=high.bikeh.fr.start, colour="orange",size=2) +
  geom_text(data=peak.bikeh.fr.start, label=peak.bikeh.fr.start$hours.start, 
            vjust= -1,col="gray",size=3) +
  facet_grid(~ days.start) +
  ylab("bike count")


# number of bikes per hour per time periods
bikeper.fr.start <- aggregate(bikeid ~ pernum.start+days.start, data = ef.sel, FUN = function(x){NROW(unique(x))})
head(bikeper.fr.start)

peak.bikeper.fr.start <- bikeper.fr.start[which(bikeper.fr.start$bikeid >= 
                                              quantile(bikeper.fr.start$bikeid,0.95)),]

high.bikeper.fr.start <- bikeper.fr.start[which(bikeper.fr.start$bikeid < 
                                              quantile(bikeper.fr.start$bikeid,0.95) &
                                              bikeper.fr.start$bikeid >= 
                                              quantile(bikeper.fr.start$bikeid,0.75)),]


par(mfrow=c(1,1))
ggplot(data=bikeper.fr.start, aes(x=pernum.start,y=bikeid,group=1)) + 

  geom_line(col="gray")+
  geom_point(col="gray")+
  geom_point(data=peak.bikeper.fr.start, colour="red",size=3) +
  geom_point(data=high.bikeper.fr.start, colour="orange",size=2) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+

  facet_grid(~ days.start)+
  ylab("bike count")

# number of bikes per hour per work day / week-end

bikew.fr.start <- aggregate(bikeid ~ hours.start+workday.start, data = ef.sel, FUN = function(x){NROW(unique(x))})
head(bikew.fr.start)

peak.bikew.fr.start <- bikew.fr.start[which(bikew.fr.start$bikeid >= 
                                              quantile(bikew.fr.start$bikeid,0.95)),]

high.bikew.fr.start <- bikew.fr.start[which(bikew.fr.start$bikeid < 
                                              quantile(bikew.fr.start$bikeid,0.95) &
                                              bikew.fr.start$bikeid >= 
                                              quantile(bikew.fr.start$bikeid,0.75)),]


lbs.w <- c("0" = "week-end", "1" = "working day")

par(mfrow=c(1,1))

ggplot(data=bikew.fr.start, aes(x=hours.start,y=bikeid,group=1)) + 

  geom_line(col="gray")+
  geom_point(col="gray")+
  geom_point(data=peak.bikew.fr.start, colour="red",size=3) +
  geom_point(data=high.bikew.fr.start, colour="orange",size=2) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+

  facet_grid(~ workday.start, labeller=labeller(workday.start = lbs.w))+
  ylab("bike count")

# number of bikes per time periods per work day / week-end

bikewper.fr.start <- aggregate(bikeid ~ pernum.start+workday.start, data = ef.sel, FUN = function(x){NROW(unique(x))})
head(bikewper.fr.start)

peak.bikewper.fr.start <- bikewper.fr.start[which(bikewper.fr.start$bikeid >= 
                                              quantile(bikewper.fr.start$bikeid,0.95)),]

high.bikewper.fr.start <- bikewper.fr.start[which(bikewper.fr.start$bikeid < 
                                              quantile(bikewper.fr.start$bikeid,0.95) &
                                              bikewper.fr.start$bikeid >= 
                                              quantile(bikewper.fr.start$bikeid,0.75)),]


lbs.wper <- c("0" = "week-end", "1" = "working day")


par(mfrow=c(1,1))
ggplot(data=bikewper.fr.start, aes(x=pernum.start,y=bikeid,group=1)) + 

  geom_line(col="gray")+
  geom_point(col="gray")+
  geom_point(data=peak.bikewper.fr.start, colour="red",size=3) +
  geom_point(data=high.bikewper.fr.start, colour="orange",size=2) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+

  facet_grid(~ workday.start, labeller=labeller(workday.start = lbs.wper))+
  ylab("bike count")

#///start stations popularity and neighborhood
# create a list (not appropriate to use here aggregation) with start station names as elements
# store the bike count (as "unique(...$bikeid)") and GPS coordinates


st.bike.l <- list()
st <- unique(ef.sel$start_station_name)
# str(st)

for (s in st){
  st.bike.l[[s]] <- list(ef.sel[which(ef.sel$start_station_name == s),"bikeid"],
                         unique(ef.sel[which(ef.sel$start_station_name == s),"start_station_longitude"]),
                         unique(ef.sel[which(ef.sel$start_station_name == s),"start_station_latitude"]),
                         unique(ef.sel[which(ef.sel$start_station_name == s),"end_station_name"]))
}


pop.st <- data.frame(lon=rep(0,length(names(st.bike.l))),
                     lat=rep(0,length(names(st.bike.l))),
                     bikecount=rep(0,length(names(st.bike.l))),
                     endstations=rep(0,length(names(st.bike.l))))
rownames(pop.st) <- names(st.bike.l)


for (l in names(st.bike.l)){
  pop.st[l,"lon"] <- st.bike.l[[l]][[2]]
  pop.st[l,"lat"] <- st.bike.l[[l]][[3]]
  pop.st[l,"bikecount"] <- length(st.bike.l[[l]][[1]])
  pop.st[l,"endstations"] <- length(st.bike.l[[l]][[4]])
}

head(pop.st[order(- pop.st$bikecount),])
tail(pop.st[order(- pop.st$bikecount),])


# correlation of bike number and connected end stations number (expected)

par(mfrow=c(1,1))

plot(log2(pop.st$endstations+1),log2(pop.st$bikecount+1),col="gray",pch=18,xlab="no. of connected end stations (log2)",
     ylab="no. of bikes (log2)")
abline(coef(lm(log2(pop.st$bikecount+1) ~ log2(pop.st$endstations+1))),col="cornflowerblue")
mtext(paste("r.sq= ",round(summary(lm(log2(pop.st$bikecount+1) ~ log2(pop.st$endstations+1)))$r.squared,2)),3,col="cornflowerblue",cex=0.75)


# plot start tations (size and color mapped onto bike counts log2) on map

#quantile(log2(pop.st$bikecount+1),c(0.75,0.95))
                                           

pop.st$plotsize <- sapply(log2(pop.st$bikecount+1), function (s) 
                                            ifelse(s <= quantile(log2(pop.st$bikecount+1),0.75),2,
                                              ifelse(s <= quantile(log2(pop.st$bikecount+1),0.95),5,8)))
pop.st$plotcol <- sapply(log2(pop.st$bikecount+1), function (s) 
                                            ifelse(s <= quantile(log2(pop.st$bikecount+1),0.75),"cornflowerblue",
                                              ifelse(s <= quantile(log2(pop.st$bikecount+1),0.95),"orange","red")))

head(pop.st)

par(mfrow=c(1,1))

NYmap <- qmap('newyork', zoom = 12,color = 'bw')

NYmap + 
  geom_point(aes(x = lon, y = lat),
                 size=pop.st$plotsize,
                 colour=pop.st$plotcol,
             pch=20,alpha=0.5,
             data = pop.st)

```


# Machine learning with rpart and 10-fold cross-validation of predictions and random forests

```{r,eval=FALSE,echo=TRUE,include=TRUE}
# data formatting
# one model per station 
# create new predictor: lag1 (shifed 1h behind) stations (all except the current station)

stations.l.test <- vector("list",length(strsplit(levels(ef.param.sel$start_station_name),",")))
names(stations.l.test) <- strsplit(levels(ef.param.sel$start_station_name),",")

shift.l.test <- vector("list",length(strsplit(levels(ef.param.sel$start_station_name),",")))
names(shift.l.test) <- strsplit(levels(ef.param.sel$start_station_name),",")

# test on random sampled stations

# set.seed(909)
# names(...)[sample(50)]



for (s in names(stations.l.test)){

  tst.set <- ef.param.sel[which(ef.param.sel$start_station_name == s),]
  bike.count.y0 <- aggregate(bikeid ~ nearestH.start, data = tst.set, FUN = function(x){NROW(unique(x))})
  tr.train.df <- bike.count.y0
  tr.train.df$bikecount.s <- log2(tr.train.df$bikeid+1)
  
  timerange <- data.frame(nearestH.start=unique(ef.param.sel$nearestH.start))
  tr.train.df.fulltime <- merge(timerange,tr.train.df,by="nearestH.start",all.x = TRUE)
  tr.train.df.fulltime$bikecount.s[is.na(tr.train.df.fulltime$bikecount.s)] <- 0
  
  #shift ahead the bike counts for the current station
  tr.train.df.fulltime$bikeidsh <- shift(tr.train.df.fulltime$bikecount.s,1,type = "lead")
  tr.train.df.fulltime$bikeidsh[is.na(tr.train.df.fulltime$bikeidsh)] <- 0
  
 
  tr.train.df.fulltime$bikeDclass <- ifelse(tr.train.df.fulltime$bikeidsh == 0,"empty",
                                            ifelse(tr.train.df.fulltime$bikeidsh <=  quantile(tr.train.df.fulltime$bikeidsh,0.05),"low",
                                                   ifelse(tr.train.df.fulltime$bikeidsh <= quantile(tr.train.df.fulltime$bikeidsh,0.5),"medium",
                                                          ifelse(tr.train.df.fulltime$bikeidsh <= quantile(tr.train.df.fulltime$bikeidsh,0.95),"high","peak"))))
  
  tr.train.df.fulltime$bikeDclass <- as.factor(tr.train.df.fulltime$bikeDclass)

  tr.train.df.fulltime$days.start.s <- wday(tr.train.df.fulltime$nearestH.start)
  tr.train.df.fulltime$hours.start.s <- hour(tr.train.df.fulltime$nearestH.start)
  tr.train.df.fulltime$workday.start <- as.numeric(isWeekday(tr.train.df.fulltime$nearestH.start))
  
  tr.train.df.fulltime$pernum.start.s <- cut(tr.train.df.fulltime$hours.start.s, c(-Inf, 7, 10, 12, 17, Inf),
                                             labels=c("night", "morning","noon","afternoon","evening"))
  
  
  tr.train.df.fulltime$workday.start <- as.factor(tr.train.df.fulltime$workday.start)
  
  tr.train.df.fulltime$days.start.s <- as.factor(tr.train.df.fulltime$days.start.s)
  tr.train.df.fulltime$hours.start.s <- as.factor(tr.train.df.fulltime$hours.start.s)

  stations.l.test[[s]] <- tr.train.df.fulltime
  
  # shift by 1h (lag) the bike counts 
  
  shift.l.test[[s]] <- data.frame(tr.train.df.fulltime$nearestH.start,
                             shift(tr.train.df.fulltime$bikecount.s,1))
  shift.l.test[[s]][,2][is.na(shift.l.test[[s]][,2])] <- 0
  
  colnames(shift.l.test[[s]]) <- c("nearestH.start",paste0("lag_",s))

  shift.l.test[[s]]$shifted <- ifelse(shift.l.test[[s]][,2] == 0,"empty",
                                          ifelse(shift.l.test[[s]][,2] <= quantile(shift.l.test[[s]][,2],0.25),"low",
                                                 ifelse(shift.l.test[[s]][,2] <= quantile(shift.l.test[[s]][,2],0.5),"medium",
                                                        ifelse(shift.l.test[[s]][,2] <= quantile(shift.l.test[[s]][,2],0.95),"high","peak"))))

  shift.l.test[[s]]$shifted <- as.factor(shift.l.test[[s]]$shifted)  
  shift.l.test[[s]][,2] <- NULL
  colnames(shift.l.test[[s]]) <- c("nearestH.start",paste0("lagclass_",s))
}

# for each ML formatted dataset (one per start station), add the shifted bike counts of the remaining stations

for (s in 1:length(stations.l.test)){  
  cat("processing item: ",s,"\n")
  shift.m <- Reduce(function(x,y) merge(x,y, all=T,by.x='nearestH.start',by.y='nearestH.start'),
                             shift.l.test[-s], 
                             accumulate=F)

  stations.l.test[[s]] <- cbind(stations.l.test[[s]],shift.m)

}


# save(stations.l.test,file="E://eFeel//RandstationsML")
# save(stations.l.test,file="/Users/Diana/eFeel/stationsML")

# ML 

# source("http://stat.ethz.ch/Teaching/WBL/Source-WBL-5/03.RCodes/dm-serie2.R")
CVtest <-function(fitfn, predfn, data, k = 10, verbose=TRUE, ...){
  n <- nrow(data)
  stopifnot(is.numeric(n), n >= 1, 1 <= k, k <= n)
  ii <- sample(n)
  res <- numeric(n)
  j1 <- 1                          ## Start des ersten Blocks
  if(verbose) cat("fold ")
  for (i in 1:k) {
    j2 <- (i*n) %/% k              ## Ende des i-ten Blocks
    j <- ii[j1:j2]                 ## Indizes der Test-Beob im Fold i
    fitted.model <- fitfn(data = data[-j,], ...)
    if(verbose) { cat(i, ""); flush.console() }
    res[j] <- predfn(fitted.model, newdata = data[j,])
    j1 <- j2 + 1                   ## Start des (i+1)-ten Blocks
  }; if(verbose) cat("\n")
  res
}

# adjust the "con" at return 
con.dcs <-function(...){
  tab <- table(...)
  t0 <- tab ; diag(t0) <- 0
  cv.er <- round(100*sum(t0)/length(list(...)[[1]]), 2)
  return(list(CV10.cm=tab,CV10.er=cv.er))
}

set.seed(909)

stationMod.l <- vector("list",length(names(stations.l.test)))
names(stationMod.l) <- names(stations.l.test)


for(m in names(stations.l.test)){
    cat("processing item: ",m,"\n")
    tst.d <- stations.l.test[[m]]
    pred <- colnames(tst.d)[grep("^lag",colnames(tst.d))] 
    pred <- c("bikeDclass","workday.start", "days.start.s","pernum.start.s",pred)
    
    res.rpart <- CVtest(function(...) rpart(bikeDclass ~ ., ...),
                    function(obj, ...) predict(obj, type = "class", ...),
                    data = tst.d[,pred])

    msc <- con.dcs(true = tst.d$bikeDclass, "CV-predicted" = res.rpart)

    rp <- rpart(bikeDclass ~ ., data=tst.d[,pred],
              control=rpart.control(cp=0.0, minsplit=5, xval=nrow(tst.d[,pred])))

    rf <- ranger(bikeDclass ~ ., data = tst.d[,pred], importance = "permutation")
    
    stationMod.l[[m]] <- list(msc,rp,rf)
    names(stationMod.l[[m]]) <- c("CV10rpart_MSC","rpart","rf")
}


# save(stationMod.l,file="E://eFeel//RandstationMod")
# save(stationMod.l,file="/Users/Diana/eFeel/stationMod")

```

# ML evaluation

```{r,eval=TRUE,echo=TRUE}

# load ML formatted list and ML models for each station
load(file="C://Users//Diana//Desktop//eFeel//stationsML")
load(file="C://Users//Diana//Desktop//eFeel//stationMod")

#/// model performance overview
# distribution of the CV10 error rate (rpart) and OOB error rate (radom forests)

CV10.er <- sapply(stationMod.l, function(x) x[[1]]$CV10.er)
str(CV10.er)
rf.er <- sapply(stationMod.l, function(x) x[[3]]$prediction.error*100)
str(rf.er)

par(mfrow=c(1,2))
hist(CV10.er,col="gray",main="rpart CV10 prediction error rate distribution",cex.main=0.75)
abline(v=quantile(CV10.er,0.5),col="cornflowerblue",lwd=2)
mtext(paste0("q5%: ",round(quantile(CV10.er,0.5))),at=quantile(CV10.er,0.5),side=3,col="cornflowerblue",cex=0.5)
hist(rf.er,col="gray",main="ranger OOB prediction error rate distribution",cex.main=0.75)
abline(v=quantile(rf.er,0.5),col="cornflowerblue",lwd=2)
mtext(paste0("q5%: ",round(quantile(rf.er,0.5))),at=quantile(rf.er,0.5),side=3,col="cornflowerblue",cex=0.5)
par(mfrow=c(1,1))

# how are "peak" and "high" bike demand predicted with CV10 rpart
maxpeak <- sapply(stationMod.l, function(x) rowSums(x[[1]]$CV10.cm)["peak"])
maxpeak[which.max(maxpeak)]

stationMod.l[["11_Ave_&_W_27_St"]][[1]]

maxhigh <- sapply(stationMod.l, function(x) rowSums(x[[1]]$CV10.cm)["high"])
maxhigh[which.max(maxhigh)]

stationMod.l[["E_72_St_&_Park_Ave"]][[1]]


# top3 most important predictors (ranger)

featureImportance <- data.frame(Feature=names(importance(stationMod.l[[1]][[3]])),Importance=importance(stationMod.l[[1]][[3]]))
a=featureImportance[order(- featureImportance$Importance),][1:10,]

rf.fi <- list()
for (l in names(stationMod.l)){
  fI <- data.frame(Feature=names(importance(stationMod.l[[l]][[3]])),Importance=importance(stationMod.l[[l]][[3]]))
  rf.fi[[l]] <- fI[order(- fI$Importance),][1:10,]
}

rf.fi.df <- do.call("rbind",rf.fi)
rownames(rf.fi.df) <- NULL
head(rf.fi.df)

pr.fr <-as.data.frame(table(rf.fi.df$Feature))
head(pr.fr[order(- pr.fr$Freq),],10)

par(mfrow=c(1,2))
hist(rf.fi.df[which(rf.fi.df$Feature == "lagclass_E_47_St_&_Park_Ave"),"Importance"],main="lagclass_E_47_St_&_Park_Ave importance ranger",cex.main=0.75,xlab="Importance")
abline(v=mean(rf.fi.df$Importance),col="cornflowerblue",lwd=2)
mtext(paste0("mean predictor importance:",round(mean(rf.fi.df$Importance),3)),cex=0.5,side=3,at=mean(rf.fi.df$Importance),col="cornflowerblue")



hist(rf.fi.df[which(rf.fi.df$Feature == "pernum.start.s"),"Importance"],main="pernum.start.s importance ranger",cex.main=0.75,xlab="Importance")
abline(v=mean(rf.fi.df$Importance),col="cornflowerblue",lwd=2)
mtext(paste0("mean predictor importance:",round(mean(rf.fi.df$Importance),3)),cex=0.5,side=3,at=mean(rf.fi.df$Importance),col="cornflowerblue")
par(mfrow=c(1,1))
```

# top 3 popular stations (ML summaries and plots)

```{r,eval=TRUE,echo=TRUE}
head(pop.st)


pop.st$plotsizesel <- sapply(log2(pop.st$bikecount+1), function (s) 
                                            ifelse(s <= quantile(log2(pop.st$bikecount+1),0.75),8,
                                              ifelse(s <= quantile(log2(pop.st$bikecount+1),0.95),15,25)))

```
## No.1 start station (i.e. highest bike count): Pershing_Square_North

```{r,eval=TRUE,echo=TRUE}

pop.st[order(- pop.st$bikecount),][1:3,]
p.st.ML <- stationMod.l[rownames(pop.st[order(- pop.st$bikecount),][1:3,])]
summary(p.st.ML)

cat("No1 popular station: ",names(p.st.ML)[1],"CV10 rpart","\n")
p.st.ML[[1]][[1]]

rp.p <- p.st.ML[[1]][[2]]
plotcp(rp.p)
printcp(rp.p)

#prp(rp, type = 4, extra = 1,varlen = 0,clip.right.labs = F,box.palette = "auto")

rp.p.pr <- prune.rpart(rp.p, cp=0.0276382)
prp(rp.p.pr, type = 4, extra = 1,varlen = 0,clip.right.labs = F,box.palette = "auto")


cat("No1 popular station: ",names(p.st.ML)[1],"ranger","\n")
rf.p <- p.st.ML[[1]][[3]]
featureImportance <- data.frame(Feature=names(importance(rf.p)),Importance=importance(rf.p))
featureImportance <- featureImportance[order(- featureImportance$Importance),][1:10,]

p <- ggplot(featureImportance, aes(x=reorder(Feature, Importance), y=Importance)) +
  geom_bar(stat="identity", fill="cornflowerblue") +
  coord_flip() + 
  theme_light(base_size=14) +
  xlab("Importance") +
  ylab("") + 
  ggtitle("Random Forest Feature Importance\n") +
  theme(plot.title=element_text(size=12))

p

# is lag station a neighbor?

# plot start tations (size and color mapped onto bike counts log2) on map

#quantile(log2(pop.st$bikecount+1),c(0.75,0.95))

stn <- c(ps="Pershing_Square_North",rp1="E_47_St_&_Park_Ave",rng="South_End_Ave_&_Liberty_St")

pop.st.sel <- pop.st[stn,]

pop.st.sel$endlon <- pop.st.sel[1,"lon"]
pop.st.sel$endlat <- pop.st.sel[1,"lat"]

pop.st.sel
```

```{r, out.width = '\\maxwidth'}
par(mfrow=c(1,1))

NYmap <- qmap('newyork', zoom = 12,color = 'bw')


NYmap + 
  geom_point(aes(x = lon, y = lat),
                 size=pop.st.sel$plotsizesel,
                 colour=pop.st.sel$plotcol,
             pch=20,alpha=0.55,
             data = pop.st.sel)+

  geom_text(data=pop.st.sel, col="blue",label=stn, 
            vjust= -1,hjust=1,size=3) +

  geom_segment(aes(x = lon, y = lat,
                       xend = endlon, yend = endlat),
                   colour="blue",
                   alpha=0.35,lwd=2,
                   pop.st.sel[-1,],arrow=arrow(type = "closed")) +
  geom_text(data = pop.st.sel[-1,], aes(x = lon, y = lat), hjust=-1.15,label = names(stn)[-1],col="magenta")

 



```
## No.2 start station (i.e. highest bike count): W_21_St_&_6_Ave

```{r,eval=TRUE,echo=TRUE}

cat("No2 popular station: ",names(p.st.ML)[2],"CV10 rpart","\n")
p.st.ML[[2]][[1]]

rp.p <- p.st.ML[[2]][[2]]
plotcp(rp.p)
printcp(rp.p)

#prp(rp, type = 4, extra = 1,varlen = 0,clip.right.labs = F,box.palette = "auto")

rp.p.pr <- prune.rpart(rp.p, cp=0.0236908)
prp(rp.p.pr, type = 4, extra = 1,varlen = 0,clip.right.labs = F,box.palette = "auto")


cat("No2 popular station: ",names(p.st.ML)[2],"ranger","\n")
rf.p <- p.st.ML[[2]][[3]]
featureImportance <- data.frame(Feature=names(importance(rf.p)),Importance=importance(rf.p))
featureImportance <- featureImportance[order(- featureImportance$Importance),][1:10,]

p <- ggplot(featureImportance, aes(x=reorder(Feature, Importance), y=Importance)) +
  geom_bar(stat="identity", fill="cornflowerblue") +
  coord_flip() + 
  theme_light(base_size=14) +
  xlab("Importance") +
  ylab("") + 
  ggtitle("Random Forest Feature Importance\n") +
  theme(plot.title=element_text(size=12))

p

# is lag station a neighbor?

# plot start tations (size and color mapped onto bike counts log2) on map

#quantile(log2(pop.st$bikecount+1),c(0.75,0.95))


stn <- c(ps="W_21_St_&_6_Ave",rp1="E_47_St_&_Park_Ave",rp2="Henry_St_&_Grand_St",rng="South_End_Ave_&_Liberty_St")

pop.st.sel <- pop.st[stn,]

pop.st.sel$endlon <- pop.st.sel[1,"lon"]
pop.st.sel$endlat <- pop.st.sel[1,"lat"]

pop.st.sel

```
```{r, out.width = '\\maxwidth'}
par(mfrow=c(1,1))

NYmap <- qmap('newyork', zoom = 12,color = 'bw')


NYmap + 
  geom_point(aes(x = lon, y = lat),
                 size=pop.st.sel$plotsizesel,
                 colour=pop.st.sel$plotcol,
             pch=20,alpha=0.55,
             data = pop.st.sel)+

  geom_text(data=pop.st.sel, col="blue",label=stn, 
            vjust= -1,hjust=1,size=3) +

  geom_segment(aes(x = lon, y = lat,
                       xend = endlon, yend = endlat),
                   colour="blue",
                   alpha=0.35,lwd=2,
                   pop.st.sel[-1,],arrow=arrow(type = "closed")) +
  geom_text(data = pop.st.sel[-1,], aes(x = lon, y = lat), hjust=-1.15,label = names(stn)[-1],col="magenta")

 


```
## No.3 start station (i.e. highest bike count): Lafayette_St_&_E_8_St

```{r,eval=TRUE,echo=TRUE}

cat("No3 popular station: ",names(p.st.ML)[3],"CV10 rpart","\n")
p.st.ML[[3]][[1]]

rp.p <- p.st.ML[[3]][[2]]
plotcp(rp.p)
printcp(rp.p)

#prp(rp, type = 4, extra = 1,varlen = 0,clip.right.labs = F,box.palette = "auto")

rp.p.pr <- prune.rpart(rp.p, cp=0.0284238)
prp(rp.p.pr, type = 4, extra = 1,varlen = 0,clip.right.labs = F,box.palette = "auto")

cat("No3 popular station: ",names(p.st.ML)[3],"ranger","\n")
rf.p <- p.st.ML[[3]][[3]]
featureImportance <- data.frame(Feature=names(importance(rf.p)),Importance=importance(rf.p))
featureImportance <- featureImportance[order(- featureImportance$Importance),][1:10,]

p <- ggplot(featureImportance, aes(x=reorder(Feature, Importance), y=Importance)) +
  geom_bar(stat="identity", fill="cornflowerblue") +
  coord_flip() + 
  theme_light(base_size=14) +
  xlab("Importance") +
  ylab("") + 
  ggtitle("Random Forest Feature Importance\n") +
  theme(plot.title=element_text(size=12))

p

# is lag station a neighbor?

stn <- c(ps="Lafayette_St_&_E_8_St",rp1.rng="E_47_St_&_Park_Ave",rp2="E_24_St_&_Park_Ave_S")

pop.st.sel <- pop.st[stn,]

pop.st.sel$endlon <- pop.st.sel[1,"lon"]
pop.st.sel$endlat <- pop.st.sel[1,"lat"]

pop.st.sel

```

```{r, out.width = '\\maxwidth'}

par(mfrow=c(1,1))

NYmap <- qmap('newyork', zoom = 12,color = 'bw')

NYmap + 
  geom_point(aes(x = lon, y = lat),
                 size=pop.st.sel$plotsizesel,
                 colour=pop.st.sel$plotcol,
             pch=20,alpha=0.55,
             data = pop.st.sel)+

  geom_text(data=pop.st.sel, col="blue",label=stn, 
            vjust= -1,hjust=1,size=3) +

  geom_segment(aes(x = lon, y = lat,
                       xend = endlon, yend = endlat),
                   colour="blue",
                   alpha=0.35,lwd=2,
                   pop.st.sel[-1,],arrow=arrow(type = "closed")) +
  geom_text(data = pop.st.sel[-1,], aes(x = lon, y = lat), hjust=-1.15,label = names(stn)[-1],col="magenta")

 

```
# R session info

```{r,eval=TRUE,echo=TRUE}

sessionInfo()


```